---
title: "2025_09_tetra_plasmid"
format: html
editor: visual
---

## Tetranucleotide similarity analysis

9/19/2025

Took the 108 CAMI2 plasmids and ran them through anvi-compute-genome-similarity –method TETRA to create a tetranucleotide similarity matrix between each of the 108 plasmids

```{r}
tetra <- read.csv("TETRA_correlations.txt", header=TRUE, row.names=1, sep="\t")
```

You can add options to executable code like this

```{r}
library(MASS)
dist_matrix <- as.dist(1 - tetra)
```

```{r}
mds_result <- isoMDS(dist_matrix, k = 2)

mds_coords <- data.frame(
  Plasmid = rownames(tetra),
  MDS1 = mds_result$points[,1],
  MDS2 = mds_result$points[,2]
)
```

```{r}
anvio_metadata <- read.csv("plasmid_CE_metadata_anvio.txt", sep="\t", header=TRUE
```

```{r}
# merge the MSD table with the metadata

mds_with_metadata <- merge(mds_coords, anvio_metadata, by.x="Plasmid", by.y="genome_id", all.x=TRUE)
```

```{r}
library(ggplot2)
library(dplyr)

# Calculate group centroids
#centroids <- mds_with_metadata %>%
#  group_by(megahit_detection) %>% 
#  summarise(
#    centroid_MDS1 = mean(MDS1, na.rm = TRUE),
#    centroid_MDS2 = mean(MDS2, na.rm = TRUE),
#    .groups = 'drop'
#  )


ggplot(mds_with_metadata, aes(x = MDS1, y = MDS2, color=megahit_detection, shape = megahit_detection)) +
  geom_point(size = 3, alpha = 0.7) +
    #geom_point(data = centroids, 
             #aes(x = centroid_MDS1, y = centroid_MDS2, fill = megahit_detection),
             #shape = 23, size = 6, color = "black", stroke = 1) + 
  labs(title = "MDS of Plasmids Based on Tetranucleotide Similarity",
       x = "MDS Dimension 1", y = "MDS Dimension 2") +
  theme_minimal() + theme(legend.title = element_text(size=12),
                          plot.title = element_text(hjust=0.5))
```

```{r}
# filter out the one blank - one blank CE for mpspades 
library(dplyr)
print(nrow(mds_with_metadata))
mds_with_metadata_mpspades <- mds_with_metadata %>%
  filter(mpspades_detection != "")
print(nrow(mds_with_metadata_mpspades))
```

```{r}
ggplot(mds_with_metadata_mpspades, aes(x = MDS1, y = MDS2, color=mpspades_detection, shape = mpspades_detection)) +
  geom_point(size = 3, alpha = 0.7) +
    #geom_point(data = centroids, 
             #aes(x = centroid_MDS1, y = centroid_MDS2, fill = megahit_detection),
             #shape = 23, size = 6, color = "black", stroke = 1) + 
  labs(title = "MDS of Plasmids Based on Tetranucleotide Similarity",
       x = "MDS Dimension 1", y = "MDS Dimension 2") +
  theme_minimal() + theme(legend.title = element_text(size=12),
                          plot.title = element_text(hjust=0.5))
```

```{r}
library(pairwiseAdonis)

mds_dist_mpspades <- dist(mds_with_metadata_mpspades[, c("MDS1", "MDS2")])
pairwise_result_mpspades <- pairwise.adonis(mds_dist_mpspades, 
                                   mds_with_metadata_mpspades$mpspades_detection,
                                   perm = 999)
print(pairwise_result_mpspades)

mds_dist_megahit <- dist(mds_with_metadata[, c("MDS1", "MDS2")])
pairwise_result_megahit <- pairwise.adonis(mds_dist_megahit, 
                                   mds_with_metadata$megahit_detection,
                                   perm = 999)
print(pairwise_result_megahit)

```

mpspades:

detected vs. undetected_genomad_plasx: F = 6.01, p adj = 0.015

detected vs. undetected_genomad: F = 0.66, p adj = 1.00

undetected_genomad_plasx vs. undetected_genomad: F = 1.44, p adj = 0.672

megahit:

detected vs. undetected_genomad_plasx: F = 9.68, p adj = 0.003

detected vs. undetected_genomad: F = 1.02, p adj = 1.000

undetected_genomad_plasx vs. undetected_genomad: F = 5.08, p adj = 0.024
